---
title: "Biostat 203B Homework 4"
subtitle: Due Mar 18 @ 11:59PM
author: Yenlin Lai
output:
  # ioslides_presentation: default
  html_document:
    toc: true
    toc_depth: 4
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

Display machine information:
```{r}
sessionInfo()
```
Load database libraries and the tidyverse frontend:
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(miceRanger))
```

## Q1. Missing data

Through the Shiny app developed in HW3, we observe abundant missing values in the MIMIC-IV ICU cohort we created. In this question, we use multiple imputation to obtain a data set without missing values.

0. Read following tutorials on the R package miceRanger for imputation: <https://github.com/farrellday/miceRanger>, <https://cran.r-project.org/web/packages/miceRanger/vignettes/miceAlgorithm.html>.

    A more thorough book treatment of the practical imputation strategies is the book [*_Flexible Imputation of Missing Data_*](https://stefvanbuuren.name/fimd/) by Stef van Buuren. 

1. Explain the jargon MCAR, MAR, and MNAR.  
**Solution:**  
MCAR: Missing completely at random. The probability of being missing is the same for all cases, meaning that causes of the missing data are unrelated to the data.  
MAR: Missing at random. The probability of being missing is the same only within groups defined by the observed data. MAR is more general and more realistic than MCAR. Modern missing data methods generally start from the MAR assumption.  
MNAR: Missing not at random. The probability of being missing varies for reasons that are unknown to us.

2. Explain in a couple of sentences how the Multiple Imputation by Chained Equations (MICE) work.  
**Solution:**
MICE is a multiple imputation method used to replace missing data values in a data set under certain assumptions about the data missingness mechanism (e.g., the data are missing at random, the data are missing completely at random).  
The procedure ‘fills in’ (imputes) missing data in a dataset through an iterative series of predictive models. In each iteration, each specified variable in the dataset is imputed using the other variables in the dataset. These iterations should be run until it appears that convergence has been met.  
This process is continued until all specified variables have been imputed. Additional iterations can be run if it appears that the average imputed values have not converged, although no more than 5 iterations are usually necessary.

3. Perform a data quality check of the ICU stays data. Discard variables with substantial missingness, say >5000 `NA`s. Replace apparent data entry errors by `NA`s.  
**Solution:**
```{r}
icu_cohort <- readRDS("icu_cohort.rds")

#decide which variables have more than 5000 NAs
na_count <- icu_cohort %>% 
  sapply(function(y) sum(is.na(y)))
na_count %>% 
  data.frame() %>% 
  filter(. > 5000)

#create a data without variables with substantial missingness
icu_cohort_na <- 
  icu_cohort %>% 
  select(-c(deathtime, edregtime, edouttime, dod, magnesium, potassium,
            sodium, white_blood_cell_count, bicarbonate, calcium, chloride, 
            creatinine, glucose, body_temperature_in_Fahrenheit, 
            systolic_non_invasive_blood_pressure, 
            mean_non_invasive_blood_pressure, respiratory_rate, heart_rate))

#define outliers
#there is only one continuous variable with NAs
outliers <- boxplot(icu_cohort_na$hematocrit, plot = FALSE)$out

#replace apparent data entry errors by NAs
icu_cohort_na[icu_cohort_na$hematocrit %in% outliers, "hematocrit"] = NA

print(icu_cohort_na, width = Inf)

na_count_na <- icu_cohort_na %>% 
  sapply(function(y) sum(is.na(y)))
na_count_na
```


4. Impute missing values by `miceRanger` (request $m=3$ data sets). This step is computational intensive. Make sure to save the imputation results as a file. Hint: Setting `max.depth=10` in the `miceRanger` function may cut some computing time.  
**Solution:**
```{r}
icu_impute <- 
  miceRanger(icu_cohort_na,
             m = 3,
             max.depth=10)
```


5. Make imputation diagnostic plots and explain what they mean.  
**Solution:**
```{r}
plotDistributions(icu_impute)
```


6. Choose one of the imputed data sets to be used in Q2. This is **not** a good idea to use just one imputed data set or to average multiple imputed data sets. Explain in a couple of sentences what the correct Multiple Imputation strategy is.  
**Solution:**

## Q2. Predicting 30-day mortality

Develop at least two analytic approaches for predicting the 30-day mortality of patients admitted to ICU using demographic information (gender, age, marital status, ethnicity), first lab measurements during ICU stay, and first vital measurements during ICU stay. For example, you can use (1) logistic regression (`glm()` function in base R or keras), (2) logistic regression with lasso penalty (glmnet or keras package), (3) random forest (randomForest package), or (4) neural network (keras package).

1. Partition data into 80% training set and 20% test set. Stratify partitioning according the 30-day mortality status.

2. Train the models using the training set.

3. Compare model prediction performance on the test set.

